|

### 2024-09-04 manuscript/paperCMSACversion.Rmd
edits and reproducing their conclusion

dat$Pos2 <- as.character(dat$Pos)
dat$Pos2[dat$Pos2 %in% c("L","R")] <- "W"
dat$Pos2 <- factor(dat$Pos2)

summ <-
  dat %>% group_by(EF) %>% summarize(
    n = n(),
    mnGP = round(mean(GP.x), 2),
    seGP = round(sd(GP.x), 2),
    mnage = round(mean(age), 2),
    seage = round(sd(age) / sqrt(n),3),
    mnwt = round(mean(Wt), 2),
    sewt = round(sd(Wt) / sqrt(n),3),
    mnht = round(mean(Ht), 2),
    seht = round(sd(Ht) / sqrt(n),3),
    mngoals = round(mean(G), 2),
    segoals = round(sd(G) / sqrt(n),3),
    mntf = round(mean(Total.Fights), 2),
    setf = round(sd(Total.Fights) / sqrt(n),3),
    mnpimgp = round(mean(PIM.GP_numeric), 2),
    sepimgp = round(sd(PIM.GP_numeric) / sqrt(n),3),
    pD = round(mean(Pos2 == "D"),3),
    pW = round(mean(Pos2 == "W"),3),
    pC = round(mean(Pos2 == "C"),3)

@@

summEP <-
  dat %>% group_by(EP) %>% summarize(
    n = n(),
    mnGP = round(mean(GP.x), 2),
    seGP = round(sd(GP.x), 2),
    mnage = round(mean(age), 2),
    seage = round(sd(age) / sqrt(n),3),
    mnwt = round(mean(Wt), 2),
    sewt = round(sd(Wt) / sqrt(n),3),
    mnht = round(mean(Ht), 2),
    seht = round(sd(Ht) / sqrt(n),3),
    mngoals = round(mean(G), 2),
    segoals = round(sd(G) / sqrt(n),3),
    mntf = round(mean(Total.Fights), 2),
    setf = round(sd(Total.Fights) / sqrt(n),3),
    mnpimgp = round(mean(PIM.GP_numeric), 2),
    sepimgp = round(sd(PIM.GP_numeric) / sqrt(n),3),
    pD = round(mean(Pos2 == "D"),3),
    pW = round(mean(Pos2 == "W"),3),
    pC = round(mean(Pos2 == "C"),3)
  )

@@

Fighting in the National Hockey League (NHL) has been an integral part of the sport since the earliest days of the league. More than just a way to express frustration, players often use fighting strategically to the benefit of the team. One of these strategies includes instigating a fight when a player from the opposite team has violated an unwritten rule of the game, which historically has led to hockey teams carrying so-called "enforcers" on their rosters. An enforcer is a member of a hockey team who, along with being a regular player, is the one tasked with fighting these opposing players. While this is not an officially defined position, the nature of needing to be able to fight any opponent generally designates the biggest and strongest players as those most likely to be enforcers.

@@

While any player on a hockey team can get into a fight, enforcers are exposed to many more fights than the typical NHL player. As a result of being exposed to additional contact, they are potentially exposed to additional head trauma beyond the amount associated with playing hockey. There are many questions about the long-term implications of this head trauma and its potential to result in negative health outcomes in the future. The ingrained nature of fighting into the game of hockey is greatly criticized and there are many active debates about its current declining trend (can we cite something that shows the trend is declining?). Should the game continue to move away from fighting in favor of younger, faster, highly-skilled players? Objectively, however, fighting has historically been a non-negligible part of the game and correctly understanding the implications this has on player health may help direct the future of the sport.

@@

<!-- Most notably, the method used to reached their most sensational finding - enforcers died on average about 10 years earlier than their non-enforcer counterparts - is statistically unsuitable. Specifically, to reach this conclusion, they use the data from the matched cohort study but then condition on death.  This effectively breaks the matching that they appropriately used to compare mortality rates, and therefore this conclusion is not statistically valid.   -->

However, we believe that some of the statistical methods used in @Popkin2023 to reach these conclusions are not the best choices given the problem that is studied.  Here, we attempt not to replicate the results of the analysis in @Popkin2023, but what we believe to be more appropriate statistical methods, specifically survival analysis techniques. This shift in perspective allows us to overcome the weakness of mortality rate, which is that it is a function of when it is calculated (i.e. In all cases, mortality rate will be 100% when all members of the original sample have died). Instead, survival analysis provides a survival curve, which can be used to analyze how the probability of survival changes over time.  In addition, we augmented the data used in @Popkin2023 to increase the sample size in the analysis and include more deceased players in the data set.

As mentioned earlier, the outcome of this study is relevant to the debates about regulations and discourse around head trauma in hockey. An accurate analysis is imperative to prevent the spread of misinformed conclusions. ... talk about our conclusions

@@

<!-- <!-- (Greg comment: After matching you CAN compare mortality rates and that comparison would be valid AT THAT GIVEN POINT when they are compared.  But a better analysis would be to use survival analysis.  The problem with mortality rate is that eventually everyone in both groups will be dead if you wait long enough and that comparison is meaningless. Survival times are much better comparison. ) -->

@@

# Data

## Data from original paper
We obtained the two data sets used in @Popkin2023: general table (n = 7,432) and death table (n = 45).  As in @Popkin2023, we considered only players who's first season was 1967 or later and played any games between the 1967-68 and the 2021-22 NHL seasons.  This leaves 6,039 players after filtering (the exact number mentioned in @Popkin2023).  We then removed 4 players with missing height, leaving us with `r nrow(dat)` (6,035) remaining observations, of which 252 are deaths, for our re-analysis.

@@

We note than in @Popkin2023, they find 183 E-P enforcers; we have only 182 in our data set with the discrepancy due the the removal of the 4 players with missing height information.

## Additional Data
While @Popkin2023 only looks at a total of 45 deceased players  (21 enforcers and 24 controls), we identified `r sum(dat$dead)` players who were included in the data used in @Popkin2023 who were deceased as of May 7, 2023 based on data from hockey-reference.com.

@@

<!-- This is what they matched on " including date of birth, total number of games played, height, weight, and position played." -->

In this same set of players, on average, enforcers in this data set tended to be older than the non-enforcers (EF: `r summ$mnage[summ$EF]` vs non-EF: `r summ$mnage[!summ$EF]` and EP: `r summEP$mnage[summEP$EP]` vs non-EP: `r summEP$mnage[!summEP$EP]`).  Additionally, as seen in Tables \ref{tab:EDAEF} and \ref{tab:EDAEP} enforcers tended to also be taller and heavier on average than non-enforcers. The means of age, weight, height and played are all significantly different between the enforcers and non-enforcer groups for both definitions of enforcers.  We also note differences some substantial differences in the positional distribution of enforcers vs non-enforcers.  Specifically, the percentage of defensive players in the enforcer vs non-enforcers is relatively close there are very few enforcers, defined with either E-F or E-P, who are listed as playing center.  This also means that the percentage of enforcers who are listed as wings (either right or left) is higher for enforcers than for non-enforcers.  The variables listed here are the same as those that were used for matching in @Popkin2023 (i.e. date of birth, total number of games played, height, weight and position).

<!-- Goals  & `r summ$mngoals[summ$EF]` (`r summ$segoals[summ$EF]`) & `r summ$mngoals[!summ$EF]` (`r summ$segoals[!summ$EF]`) & `r ifelse(round(t.test(dat$G~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$G~dat$EF)$p.value,2))`\\ -->

\begin{table}
\begin{center}
\begin{tabular}{|c|c|}
\hline
 Variable & mean (sd)\\
 \hline
 n & `r nrow(dat)`    \\
 n deceased & `r sum(dat$dead)` (`r paste0(round(100*sum(dat$dead)/nrow(dat),2),"\\%")`)    \\
 Games Played & `r  round(mean(dat$GP.x),2)` (`r round(sd(dat$GP.x),2)`)\\
 Age (Years) & `r round(mean(as.numeric(dat$age)),2)` (`r round(sd(as.numeric(dat$age)),2)`)\\
 Height (Inches) & `r round(mean(as.numeric(dat$Ht)),2)` (`r round(sd(as.numeric(dat$Ht)),2)`)\\
 Weight (Pounds) & `r mean(as.numeric(dat$Wt))` (`r sd(as.numeric(dat$Wt))`)\\
 Fights per game & `r mean(as.numeric(dat$FperG))` (`r sd(as.numeric(dat$FperG))`)\\
 Penalty Minutes per game  & `r mean(as.numeric(dat$PIM.GP_numeric))`  (`r  sd(as.numeric(dat$PIM.GP_numeric))`)\\
 \hline
\end{tabular}
\end{center}
\caption{Overall summary stats
\label{tab:EDAoverall}}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
 Variable & EF = TRUE & EF = FALSE & p-value \\
 \hline
 n & `r summ$n[summ$EF]` & `r summ$n[!summ$EF]` & \\
 Games Played & `r summ$mnGP[summ$EF]` (`r summ$seGP[summ$EF]`) & `r summ$mnGP[!summ$EF]` (`r summ$seGP[!summ$EF]`) & `r ifelse(round(t.test(dat$GP.x~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$GP.x~dat$EF)$p.value,2))`\\
 Age (years) & `r summ$mnage[summ$EF]` (`r summ$seage[summ$EF]`) & `r summ$mnage[!summ$EF]` (`r summ$seage[!summ$EF]`) & `r ifelse(round(t.test(dat$age~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EF)$p.value,2))`\\
 Weight (lbs) & `r summ$mnwt[summ$EF]` (`r summ$sewt[summ$EF]`) & `r summ$mnwt[!summ$EF]` (`r summ$sewt[!summ$EF]`) & `r ifelse(round(t.test(dat$Wt~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EF)$p.value,2))`\\
  Height (in) & `r summ$mnht[summ$EF]` (`r summ$seht[summ$EF]`) & `r summ$mnht[!summ$EF]` (`r summ$seht[!summ$EF]`) & `r ifelse(round(t.test(dat$Ht~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$Ht~dat$EF)$p.value,2))`\\
          percent D & `r paste0(100*summ$pD[summ$EF],"\\%")` & `r paste0(100*summ$pD[!summ$EF],"\\%")` & \\
    percent W & `r paste0(100*summ$pW[summ$EF],"\\%")` & `r paste0(100*summ$pW[!summ$EF],"\\%")` & \\
    percent C & `r paste0(100*summ$pC[summ$EF],"\\%")` & `r paste0(100*summ$pC[!summ$EF],"\\%")` & \\

 \hline
\end{tabular}
\end{center}
\caption{Comparing enforcers to non enforcers when enforcer is defined as E-F \label{tab:EDAEF}}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
 Variable & EP = TRUE & EP = FALSE & p-value \\
 \hline
 n & `r summEP$n[summEP$EP]` & `r summEP$n[!summEP$EP]` & \\
 Games Played & `r summEP$mnGP[summEP$EP]` (`r summEP$seGP[summEP$EP]`) & `r summEP$mnGP[!summEP$EP]` (`r summEP$seGP[!summEP$EP]`) & `r ifelse(round(t.test(dat$GP.x~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$GP.x~dat$EP)$p.value,2))`\\
 Age (years) & `r summEP$mnage[summEP$EP]` (`r summEP$seage[summEP$EP]`) & `r summEP$mnage[!summEP$EP]` (`r summEP$seage[!summEP$EP]`) & `r ifelse(round(t.test(dat$age~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EP)$p.value,2))`\\
 Weight (lbs) & `r summEP$mnwt[summEP$EP]` (`r summEP$sewt[summEP$EP]`) & `r summEP$mnwt[!summEP$EP]` (`r summEP$sewt[!summEP$EP]`) & `r ifelse(round(t.test(dat$Wt~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EP)$p.value,2))`\\
  Height (in) & `r summEP$mnht[summEP$EP]` (`r summEP$seht[summEP$EP]`) & `r summEP$mnht[!summEP$EP]` (`r summEP$seht[!summEP$EP]`) & `r ifelse(round(t.test(dat$Ht~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$Ht~dat$EP)$p.value,2))`\\
    percent D & `r paste0(100*summEP$pD[summEP$EP],"\\%")` & `r paste0(100*summEP$pD[!summEP$EP],"\\%")` & \\
    percent W & `r paste0(100*summEP$pW[summEP$EP],"\\%")` & `r paste0(100*summEP$pW[!summEP$EP],"\\%")` & \\
    percent C & `r paste0(100*summEP$pC[summEP$EP],"\\%")` & `r paste0(100*summEP$pC[!summEP$EP],"\\%")` & \\
 \hline
\end{tabular}
\end{center}

\caption{Comparing enforcers to non enforcers when enforcer is defined as E-P \label{tab:EDAEP}}
\end{table}

@@

<!-- ```{r KMEP, echo = FALSE, fig.cap="Kaplan Meier plot for EP" } -->
<!-- dat <- dat %>% mutate(northamerica = ifelse(Ctry %in% c("USA","CAN"),"North America","Other")) -->
<!-- library(ggsurvfit) -->
<!-- survfit2(Surv ~ northamerica, data = dat) %>%  -->
<!--   ggsurvfit() + -->
<!--   labs( -->
<!--     x = "Age", -->
<!--     y = "Overall survival probability" -->
<!--   ) +  -->
<!--   add_confidence_interval() -->
<!-- ``` -->

@@

# Results {#sec:results}
Table \ref{tab:EF} shows the results for the Cox proportional hazards model when enforcers were defined based on their number of career fights.  No statistically significant difference was observed in the hazard functions of enforcers versus non enforcers (p-value = 0.549). However, a significant difference is observed between enforcers and non-enforcers when the enforcer is defined based on their number of penalty minutes per game (p-value = 0.00169).  The estimated hazard ratio for E-P is 2.29, indicating a 129\% increase in the hazard for enforcers vs non-enforcers when all other variables are held constant.

Figure \@ref{fig:estsurv} shoes the estimated survival curves for enforcers versus non-enforcers (E-P) for defenders with all continuous variables held constant at mean levels.  Using this, the estimated survival rate at the age of 50 for enforcers vs non-enforcers is 96.57\% (95\% CI: 94.63\%, 98.56\%) vs 98.49\% (95\% CI: 97.98\%, 99.00\%).  For the same conditions at the age of 75, the estimated survival rate for enforcers is about 74.4\% (95\% CI: 61.4\%, 90.32\%) compared to non-enforcers estimate survival rate of 87.95\% (95\% CI: 83.09\%, 93.10\%).

@@

```{r estsurv, echo = FALSE, fig.cap = "Estimated Survival curve for EP for defenders.  All other variables are fixed at their mean level (GP = 285.8, Ht = 72.7in, Wt = 197.03, and Year of birth 1974.5."}
coxphmod_EP <- coxph(Surv ~ EP + GP.x + Ht + Wt + DOB.Year + Pos2, data = dat)
# summary(coxphmod_EP)
# library(xtable)
# xtable(summary(coxphmod_EP)$coef[,-4],digits = 5)
# cox.zph(coxphmod_EP)

@@

Unobserved Confounders: 
We further note several additional difference between the enforcers and the non-enforcer groups that were not considered in the @Popkin2023.  As an example, the number of average number of goals scores between the two groups significantly different for both definitions of enforcers.  The average number of goals scores for E-F enforcers  vs non-enforcers was `r summ$mngoals[summ$EF]` vs `r summ$mngoals[!summ$EF]` (p-value: `r ifelse(round(t.test(dat$G~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$G~dat$EF)$p.value,2))`).  For E-P enforcers, the average number of goals for enforcers was `r summEP$mngoals[summEP$EP]` vs `r summEP$mngoals[!summEP$EP]` (p-value: `r ifelse(round(t.test(dat$G~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$G~dat$EP)$p.value,2))`).

```{r echo = FALSE}
#Enforcer by country
epcountry <- dat %>% mutate(northamerica = ifelse(Ctry %in% c("USA","CAN"),"North America","Other")) %>% group_by(EP, northamerica) %>% summarize(n = n()) %>% pivot_wider(names_from = northamerica, values_from = n) %>% mutate(p = `North America`/(`North America`+ Other))

#prop.test(epcountry$`North America`,epcountry$Other+epcountry$`North America`)

#Enforcer by country
efcountry <- dat %>% mutate(northamerica = ifelse(Ctry %in% c("USA","CAN"),"North America","Other")) %>% group_by(EF, northamerica) %>% summarize(n = n()) %>% pivot_wider(names_from = northamerica, values_from = n)  %>% mutate(p = `North America`/(`North America`+ Other))

#prop.test(efcountry$`North America`,efcountry$Other+efcountry$`North America`)

#How many players are in each group
#table(dat$EP, dat$EF)

Perhaps more interestingly, we find highly significant differences in geographic region of enforcers vs non-enforcers for both definitions.  In the enforcer group `r paste0(100*round(efcountry$p[efcountry$EF],4),"\\%")` and `r paste0(100*round(epcountry$p[epcountry$EP],4),"\\%")` of those groups are from North America (i.e. United States or Canada) for E-F and E-P, respectively.  For non-enforcers, North American players make up only `r paste0(100*round(efcountry$p[!efcountry$EF],4),"\\%")` for E-F and `r paste0(100*round(epcountry$p[!epcountry$EP],4),"\\%")` for E-P.  For both definitions of enforcer, these differences are significant with p-values for both definitions less than 0.0001.

While we do not address this imbalance here, as we focus on closely following the work of @Popkin23, it is worth noting the substantial imbalance in geographic origin (i.e. North American vs outside of North America) of enforcers versus non-enforcer.

@@

>>>>

summ <-
  dat %>% group_by(EF) %>% summarize(
    n = n(),
    mnage = round(mean(age), 2),
    seage = round(sd(age) / sqrt(n),3),
    mnwt = round(mean(Wt), 2),
    sewt = round(sd(Wt) / sqrt(n),3),
    mnht = round(mean(Ht), 2),
    seht = round(sd(Ht) / sqrt(n),3),
    mngoals = round(mean(G), 2),
    segoals = round(sd(G) / sqrt(n),3),
    mntf = round(mean(Total.Fights), 2),
    setf = round(sd(Total.Fights) / sqrt(n),3),
    mnpimgp = round(mean(PIM.GP_numeric), 2),
    sepimgp = round(sd(PIM.GP_numeric) / sqrt(n),3)

@@

summEP <-
  dat %>% group_by(EP) %>% summarize(
    n = n(),
    mnage = round(mean(age), 2),
    seage = round(sd(age) / sqrt(n),3),
    mnwt = round(mean(Wt), 2),
    sewt = round(sd(Wt) / sqrt(n),3),
    mnht = round(mean(Ht), 2),
    seht = round(sd(Ht) / sqrt(n),3),
    mngoals = round(mean(G), 2),
    segoals = round(sd(G) / sqrt(n),3),
    mntf = round(mean(Total.Fights), 2),
    setf = round(sd(Total.Fights) / sqrt(n),3),
    mnpimgp = round(mean(PIM.GP_numeric), 2),
    sepimgp = round(sd(PIM.GP_numeric) / sqrt(n),3)
  )

@@

Fighting in the National Hockey League (NHL) has been an integral part of the sport since the earliest days of the league. More than just a way to express frustration, players often use fighting strategically to the benefit of the team. One of these strategies includes instigating a fight when a player from the opposite team has violated an unwritten rule of the game, which historically has led to hockey teams carrying so-called "enforcers" on their rosters. The enforcer is the teammember tasked with fighting these opposing players. While this is not an officially defined position, the nature of needing to be able to fight any opponent generally designates the biggest and strongest players as those most likely to be enforcers.

@@

While any player on a hockey team can get into a fight, enforcers are exposed to many more fights than the typical NHL player. As a result of being exposed to additional contact, they are potentially exposed to additional head trauma beyond the amount associated with playing hockey. There are many questions about the long-term implications of this head trauma and its potential to result in negative health outcomes in the future. The ingrained nature of fighting into the game of hockey is greatly criticized and there are many active debates about its current declining trend. Figure \ref{fig:enforcer_density} shows a decline in the proportion of enforcers to non-enforcers as players get younger. The question is, should the game continue to move away from contact-driven strategy in favor of younger, faster, highly-skilled players?

```{r enforcer_density, echo = FALSE, fig.cap='Density Plot of Study Cohorts by DOB'}
cohort_dat = dat %>% mutate(cohort = case_when(PIM.GP > 3 ~ 'Enforcer',
                                  Total.Fights > 50 ~ 'Enforcer',
                                  .default = 'Non-Enforcer'))

cohort_dat$cohort = as.factor(cohort_dat$cohort)

ggplot(cohort_dat, aes(x = DOB)) + geom_density(aes(fill = cohort),alpha = 0.75 )
```

@@

However, we believe that some of the methods used in @Popkin2023 to reach these conclusions are not statistically appropriate.  Most notably, the method used to reached their most sensational finding - enforcers died on average about 10 years younger than their non-enforcer counterparts - is statistically unsuitable. Specifically, to reach this conclusion, they use the data from the matched cohort study but then condition on death.  This effectively breaks the matching that they appropriately used to compare mortality rates, and therefore this conclusion is not statistically valid.

Here, we attempt not to replicate the results of the analysis in @Popkin2023, but what we believe to be more appropriate statistical analysis, specifically survival analysis methodology. This shift in perspective allows us to overcome the weakness of mortality rate, which is that it is a function of when it is calculated (i.e. In all cases, mortality rate will be 100% when all members of the original sample have died). Instead, survival analysis provides a survival curve, which can be used to analyze how the probability of survival changes over time.  In addition, we augmented the data used in @Popkin2023 to increase the sample size in the analysis and include more deceased players in the data set.

As mentioned earlier, the outcome of this study is relevant to the debates about regulations and discourse around head trauma in hockey. An accurate analysis is imperative to prevent the spread of misinformed conclusions and best inform regulatory safety practices going forward.

... talk about our conclusions

@@

<!-- <!-- (Greg comment: After matching you CAN compare mortality rates and that comparison would be valid AT THAT GIVEN POINT when they are compared.  But a better analysis would be to use survival analysis.  The problem with mortality rate is that eventually everyone in both groups will be dead if you wait long enough and that comparison is meaningless. Survival times are much better comparison. ) --> -->

@@

# Data 
## Data from orignal paper
We obtained the two data sets used in @Popkin2023: general table (n = 7,432) and death table (n = 45).  As in @Popkin2023, we considered only players who's first season began in 1967 or later and played any games between the 1967-68 and the 2021-22 NHL seasons.  This leaves 6,039 players after filtering (the exact number mentioned in @Popkin2023). We then removed 4 players with missing height, leaving us with `r nrow(dat)`  remaining observations, for our re-analysis.

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|c|}
\hline
 Variable & mean (sd)\\
 \hline
 n & `r nrow(dat)`    \\
 n deceased & `r sum(dat$dead)` (`r paste0(round(100*sum(dat$dead)/nrow(dat),2),"\\%")`)    \\
 Age (Years) & `r round(mean(as.numeric(dat$age)),2)` (`r round(sd(as.numeric(dat$age)),2)`)\\
 Height (Inches) & `r round(mean(as.numeric(dat$Ht)),2)` (`r round(sd(as.numeric(dat$Ht)),2)`)\\
 Weight (Pounds) & `r mean(as.numeric(dat$Wt))` (`r sd(as.numeric(dat$Wt))`)\\
 Fights per game & `r mean(as.numeric(dat$FperG))` (`r sd(as.numeric(dat$FperG))`)\\
 Penalty Minutes per game  & `r mean(as.numeric(dat$PIM.GP_numeric))`  (`r  sd(as.numeric(dat$PIM.GP_numeric))`)\\
 \hline
\end{tabular}
\end{center}
\caption{Overall summary stats
\label{tab:EDAoverall}}
\end{table}

## Additional Data
While @Popkin2023 only looks at a total of 45 deceased players  (21 enforcers and 24 controls), we identified `r sum(dat$dead)` players who were included in the data used in @Popkin2023 who were deceased as of May 7, 2023 based on data from hockey-reference.com. Of these additional `r sum(dat$dead) - nrow(death_JAMA)` observations of death, 2 fall in the E-F cohort, and 4 fall in the E-P cohort.

@@

Of the players included in the analysis, on average, enforcers in this data set tended to be older than the non-enforcers (EF: `r summ$mnage[summ$EF]` vs non-EF: `r summ$mnage[!summ$EF]` and EP: `r summEP$mnage[summEP$EP]` vs non-EP: `r summEP$mnage[!summEP$EP]`).  Additionally, as seen in Tables \ref{tab:EDAEF} and \ref{tab:EDAEP} enforcers tended to also be taller and heavier on average than non-enforcers. The means of age, weight, height and goals are all significantly different between the enforcers and non-enforcer groups for both definitions of enforcers.

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
 Variable & EF = TRUE & EF = FALSE & p-value \\
 \hline
 n & `r summ$n[summ$EF]` & `r summ$n[!summ$EF]` & \\
 Age (years) & `r summ$mnage[summ$EF]` (`r summ$seage[summ$EF]`) & `r summ$mnage[!summ$EF]` (`r summ$seage[!summ$EF]`) & `r ifelse(round(t.test(dat$age~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EF)$p.value,2))`\\
 Weight (lbs) & `r summ$mnwt[summ$EF]` (`r summ$sewt[summ$EF]`) & `r summ$mnwt[!summ$EF]` (`r summ$sewt[!summ$EF]`) & `r ifelse(round(t.test(dat$Wt~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EF)$p.value,2))`\\
  Height (in) & `r summ$mnht[summ$EF]` (`r summ$seht[summ$EF]`) & `r summ$mnht[!summ$EF]` (`r summ$seht[!summ$EF]`) & `r ifelse(round(t.test(dat$Ht~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$Ht~dat$EF)$p.value,2))`\\
    Goals  & `r summ$mngoals[summ$EF]` (`r summ$segoals[summ$EF]`) & `r summ$mngoals[!summ$EF]` (`r summ$segoals[!summ$EF]`) & `r ifelse(round(t.test(dat$G~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$G~dat$EF)$p.value,2))`\\
 \hline
\end{tabular}
\end{center}
\caption{Comparing enforcers to non-enforcers when enforcer is defined as E-F \label{tab:EDAEF}}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
 Variable & EP = TRUE & EP = FALSE & p-value \\
 \hline
 n & `r summEP$n[summEP$EP]` & `r summEP$n[!summEP$EP]` & \\
  GP & `r summEP$mnage[summEP$EP]` (`r summEP$seage[summEP$EP]`) & `r summEP$mnage[!summEP$EP]` (`r summEP$seage[!summEP$EP]`) & `r ifelse(round(t.test(dat$age~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EP)$p.value,2))`\\
 Age (years) & `r summEP$mnage[summEP$EP]` (`r summEP$seage[summEP$EP]`) & `r summEP$mnage[!summEP$EP]` (`r summEP$seage[!summEP$EP]`) & `r ifelse(round(t.test(dat$age~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EP)$p.value,2))`\\
 Weight (lbs) & `r summEP$mnwt[summEP$EP]` (`r summEP$sewt[summEP$EP]`) & `r summEP$mnwt[!summEP$EP]` (`r summEP$sewt[!summEP$EP]`) & `r ifelse(round(t.test(dat$Wt~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EP)$p.value,2))`\\
  Height (in) & `r summEP$mnht[summEP$EP]` (`r summEP$seht[summEP$EP]`) & `r summEP$mnht[!summEP$EP]` (`r summEP$seht[!summEP$EP]`) & `r ifelse(round(t.test(dat$Ht~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$Ht~dat$EP)$p.value,2))`\\
      Goals  & `r summEP$mngoals[summEP$EP]` (`r summEP$segoals[summEP$EP]`) & `r summEP$mngoals[!summEP$EP]` (`r summEP$segoals[!summEP$EP]`) & `r ifelse(round(t.test(dat$G~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$G~dat$EP)$p.value,2))`\\
 \hline
\end{tabular}
\end{center}

\caption{Comparing enforcers to non-enforcers when enforcer is defined as E-P \label{tab:EDAEP}}
\end{table}

@@

@@

# Results {#sec:results}
Table \ref{tab:EF} shows the results for the Cox proportional hazards model when enforcers were defined based on their number of career fights.  No statistically significant difference was observed in the hazard functions of enforcers versus non-enforcers (p-value = 0.549). However, a significant difference is observed between enforcers and non-enforcers when the enforcer is defined based on their number of penalty minutes per game (p-value = 0.00169).  The estimated hazard ratio for E-P is 2.29, indicating a 129\% increase in the hazard for enforcers vs non-enforcers when all other variables are held constant.

Figure \ref{fig:estsurv} shows the estimated survival curves for enforcers (E-P) versus non-enforcers for defenders with all continuous variables held constant at mean levels.  Using this, the estimated survival rate at the age of 50 for enforcers vs non-enforcers is 96.57\% (95\% CI: 94.63\%, 98.56\%) vs 98.49\% (95\% CI: 97.98\%, 99.00\%).  For the same conditions at the age of 75, the estimated survival rate for enforcers is about 74.4\% (95\% CI: 61.4\%, 90.32\%) compared to non-enforcers estimate survival rate of 87.95\% (95\% CI: 83.09\%, 93.10\%).  The estimated survival rate for enforcers drops an additional `r (96.57-74.4)-(98.49 - 87.95)`\% between the ages of 50 and 75 versus non-enforcers.

These results suggest that the conclusions of @Popkin2023 are overstated due to the differences in exposure time between the groups at the time of their final analysis.

join all from jama nhl data, condition on death from jama deaths and look at yob between enforcer/control
```{r}
cf = read.csv('../JAMA NHL Data/C-F-Table 1.csv')
cp = read.csv('../JAMA NHL Data/C-P-Table 1.csv')

ef = read.csv('../JAMA NHL Data/E-F-Table 1.csv')
ep = read.csv('../JAMA NHL Data/E-P-Table 1.csv')

jama_controls = dat %>% filter(Player %in% cf$Player | Player %in% cp$Player)

jama_enforcers = dat %>% filter(Player %in% ef$Player | Player %in% ep$Player)

jama_short = rbind(jama_controls, jama_enforcers)

jama_deaths = read.csv('../JAMA NHL Data/Deaths-Table 1.csv')

jama_short %>%
  filter(Player %in% jama_deaths$Enforcers) %>%
  mutate(enforcer = ifelse(EF == 1 | EP == 1, 1, 0)) %>% 
  group_by(enforcer) %>%
  summarize(mean(DOB.Year))

jama_short %>%
  mutate(deathage = as.numeric(difftime(Death.Date,
                                        Birth.Date,
                                        units = "weeks")/52.25)) %>%
  filter(Player %in% jama_deaths$Enforcers) %>%
  mutate(enforcer = ifelse(EF == 1 | EP == 1, 1, 0)) %>% 
  group_by(enforcer) %>%
  summarize(mean(deathage))

@@

```{r estsurv, echo = FALSE, fig.cap = "Estimated Survival curve for E-P for defenders.  All other variables are fixed at their mean level (GP = 285.8, Ht = 72.7in, Wt = 197.03, and Year of birth 1974.5)"}
coxphmod_EP <- coxph(Surv ~ EP + GP.x + Ht + Wt + DOB.Year + Pos2, data = dat)
# summary(coxphmod_EP)
# library(xtable)
# xtable(summary(coxphmod_EP)$coef[,-4],digits = 5)
# cox.zph(coxphmod_EP)

@@

@@

We would like to thank @Popkin2023 for their generosity and willingness to share the data provided for the above analysis.
